Sparse spectral and $p$-finite element methods on disk-slices and trapeziums I: zero boundary conditions

### 2019-04-01 papers/halfdisk/halfdisk.tex
Additions to paper

\newcommand{\half}{\frac{1}{2}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\No}{\mathbb{N}_0}
\newcommand{\Ylm}{Y^m_l}
\newcommand{\Ylmfull}{Y^m_l(\theta,\varphi)}
\newcommand{\Plm}{P^m_l}
\newcommand{\costheta}{\cos\theta}
\newcommand{\sintheta}{\sin\theta}
\newcommand{\cosphi}{\cos\varphi}
\newcommand{\sinphi}{\sin\varphi}
\newcommand{\eimphi}{e^{im\varphi}}
\newcommand{\alphalm}{\alpha^m_l}
\newcommand{\clm}{c^m_l}
\newcommand{\ctilde}{\tilde{c}^m_l}
\newcommand{\ctildemod}{\tilde{c}^{|m|}_l}
\newcommand{\chat}{\hat{c}^m_l}
\newcommand{\chatmod}{\hat{c}^{|m|}_l}
\newcommand{\ddx}{\frac{\mathrm{d}}{\mathrm{d}x}}
\newcommand{\pddx}{\frac{\partial}{\partial x}}
\newcommand{\pddy}{\frac{\partial}{\partial y}}
\newcommand{\dmdxm}{\frac{\mathrm{d}^m}{\mathrm{d}x^m}}

@@

\newcommand{\Pnk}{P_{n,k}}
\newcommand{\Pab}{P^{(a,b)}}
\newcommand{\Pnkab}{P_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\Pmjab}{P_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bigPab}{\bigP^{(a,b)}}
\newcommand{\Dnt}{D^T_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\Pii}{P^{(1,1)}}
\newcommand{\bigPii}{{\mathbb{P}^{(1,1)}}}
\newcommand{\Poo}{P^{(0,0)}}
\newcommand{\bigPoo}{{\mathbb{P}^{(0,0)}}}
\newcommand{\Poi}{P^{(0,1)}}
\newcommand{\Pio}{P^{(1,0)}}
\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}
\newcommand{\element}{\tau}
\newcommand{\refelement}{\hat{\tau}}
\newcommand{\FEset}{\mathcal{T}}
\newcommand{\bigW}{\mathbb{W}}
\newcommand{\bigWab}{\mathbb{W}^{(a,b)}}
\newcommand{\bigWii}{{\mathbb{W}^{(1,1)}}}
\newcommand{\bigQ}{\mathbb{Q}}
\newcommand{\bigQab}{\bigQ^{(a,b)}}

@@

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family\bstodo{Is this OK?}. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

We can then obtain the matrix operator for \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\bigWii(x,y) = \Wii(x,y) \: \bigPii(x,y)
\]
to coefficients in the non-weighted space
\[
\bigPii(x,y).
\]
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$. The matrix operator \(\Delta\) acting on the coefficients vector is then given by
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}

Further, we can also obtain the matrix operator for \(\Delta\) that will take us from coefficients for expansion in the space
\[
\bigPoo(x,y)
\]
to coefficients in the space
\[
\bigP^{(2,2)}(x,y).
\]
Importantly, again, this operator will have Banded-Block-Banded structure, and hence will be sparse. The matrix operator \(\Delta\) acting on the coefficients vector is then given by
\begin{align}
    \Delta := D_x^{(1,1)} \: D_x^{(0,0)} + T^{(1,2)\to(2,2)} \: T^{(0,2)\to(1,2)}  \: D_y^{(0,1)} \: D_y^{(0,0)}.
\end{align}

The sparsity of the these operators can be shown using a simple integration by parts argument. We note that the non-zero entries of the differential operators are calculated as follows.

Note that we can expand $\pddx \Pnkab = \sum_{m=0}^{n} \sum_{j=0}^m c_{m,j} \: P^{(a+1, b+1)}_{m,j}$. The coefficients of these expansion are then the entries of the relevant operator matrix. By a simple integration by parts argument, we have that the only non-zero coefficients of this expansion are when $m = n-1, n-2$, $j = k, k-2$ and $0 \le j \le m$. These are then calculated by taking inner products, i.e. 
\begin{align*}
c_{m,j} = \frac{\ip<\pddx \Pnkab, P^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}}}{\norm{P^{(a+1, b+1)}_{m,j}}^2_{W^{(a+1, b+1)}}}
\end{align*}
where 
\begin{align}
	&\ip<\pddx \Pnkab, P^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}} \nonumber \\ 
	&= \ip< x \: H_{n-k}^{(a, b+k+\half)} \prime (x), \: \rho(x)^{k+j+2} \: H_{m-j}^{(a, b+1+j+\half)} >_{w_H^{(a, b+\half)}} 
		\: \ip< P_k^{(b,b)}, \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  \nonumber \\ 
	& \quad - k \: \ip< x^2 \: H_{n-k}^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: H_{m-j}^{(a, b+1+j+\half)} >_{w_H^{(a, b+\half)}} 
		\: \ip< P_k^{(b,b)}, \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  \nonumber \\ 
	& \quad + \: \ip< x^2 \: H_{n-k}^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: H_{m-j}^{(a, b+1+j+\half)} >_{w_H^{(a, b+\half)}} 
		\: \ip< y \: P_k^{(b,b)} \prime (y), \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  
\end{align}
We can then use quadrature to calculate the 1D inner products. This is much cheaper than using the 2D quadrature rule.

Further, we can expand $\pddy \Pnkab = \sum_{m=0}^{n} \sum_{j=0}^m c_{m,j} \: P^{(a, b+1)}_{m,j}$. The coefficients of these expansion are then the entries of the relevant operator matrix. By a simple integration by parts argument, we have that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. This is then calculated by taking inner products, i.e. 
\begin{align*}
c_{n-1,k-1} = \frac{\ip<\pddx \Pnkab, P^{(a, b+1)}_{n-1,k-1} >_{W^{(a, b+1)}}}{\norm{P^{(a, b+1)}_{n-1,k-1}}^2_{W^{(a, b+1)}}}
\end{align*}
where 
\begin{align}
	\ip<\pddx \Pnkab, P^{(a, b+1)}_{n-1,k-1} >_{W^{(a, b+1)}} = \norm{H_0^{(a,b+k+\half)}}^2_{w_H^{(a, b+k+\half)}} 
		\: \ip< P_k^{(b,b)} \prime, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}}
\end{align}
where we note that $\norm{H_n^{(a,b)}}_{w_H^{(a,b)}} = \norm{H_m^{(a,b)}}_{w_H^{(a,b)}}$ $\forall m, n \in \N_0$.

@@

@@

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigPii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. In matrix-vector notation, our system hence becomes:
\begin{align}
    \Delta^{(1,1)} \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$.

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.4]{sparsityoflaplacian}
        \label{fig:sparsity}
    	\caption{A "spy" plot of the $\Delta$ operator for $\bigWii(x,y) \to \bigPii(x,y)$ showing its sparsity.}
	\centering
\end{figure}

\subsection{Inhomogenous Helmholtz}

Find \(u(x,y)\) given a function \(f(x,y)\) such that:
\begin{align}
	\begin{cases}
    		\Delta u(x,y) + k^2 u(x,y) = f(x,y) \quad \text{in } \Omega \\
		u(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
\end{align}
noting the imposition of zero Dirichlet boundary conditions on $u$.

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigPii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. In matrix-vector notation, our system hence becomes:
\begin{align}
    (\Delta^{(1,1)} + k^2 T^{(0,0)\to(1,1)} T_W^{(1,1)\to(0,0)}) \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$.

We can extend this to constant non-zero boundary conditions by simply noting that the problem 
\begin{align}
	\begin{cases}
    		\Delta u(x,y) + k^2 u(x,y) = f(x,y) \quad \text{in } \Omega \\
		u(x,y) = c \in \R \quad \text{on } \partial \Omega
	\end{cases}
\end{align}
is equivalent to letting $u = v + c$ and solving
\begin{align}
	\begin{cases}
    		\Delta v(x,y) + k^2 v(x,y) = f(x,y) - k^2 c =: g(x,y)  \quad \text{in } \Omega \\
		v(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
\end{align}

@@

\subsection{Heat Equation}
Consider
\begin{align}
	\begin{cases}
		\frac{\partial u}{\partial t}(t,x,y) = \Delta u(t,x,y) \quad \text{in } \Omega \\
		u(t, x, y) = 0 \quad \text{on } \partial \Omega \\
		u(0, x, y) = u_0(x,y) := \Wii(x,y) \: f(x,y)
	\end{cases}
\end{align}
noting the imposition of zero Dirichlet boundary conditions.

@@

Then, using the operator $\Delta$, we can solve this system for the given initial condition $\mathbf{u_0}$ by iterating for $t = 0,1,2,\dots,t_{max}$:
\begin{align}
	&T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} \: (\mathbf{u}_{t+1} - \mathbf{u}_t) = h \: \Delta^{(1,1)} \mathbf{u}_{t+1} \nonumber \\
	\iff& (T^{(0,0)\to(1,1)} \:T_W^{(1,1)\to(0,0)} - h \: \Delta^{(1,1)}) \: \mathbf{u}_{t+1} = T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} \: \mathbf{u}_t \nonumber \\
	\iff& \mathbf{u}_{t+1} = (T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} - h \: \Delta^{(1,1)})^{-1} T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} \: \mathbf{u}_t \nonumber \\
	\iff& \mathbf{\tilde{u}}_{t+1} = T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} \: (T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} - h \: \Delta^{(1,1)})^{-1} \: \mathbf{\tilde{u}}_t.
\end{align}

@@

%
\section{Future work - orthogonal polynomials on the half-spheres and spherical triangles}

>>>>

\newcommand{\half}{\frac{1}{2}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\No}{\mathbb{N}_0}
\newcommand{\Ylm}{Y^m_l}
\newcommand{\Ylmfull}{Y^m_l(\theta,\varphi)}
\newcommand{\Plm}{P^m_l}
\newcommand{\costheta}{\cos\theta}
\newcommand{\sintheta}{\sin\theta}
\newcommand{\cosphi}{\cos\varphi}
\newcommand{\sinphi}{\sin\varphi}
\newcommand{\eimphi}{e^{im\varphi}}
\newcommand{\alphalm}{\alpha^m_l}
\newcommand{\clm}{c^m_l}
\newcommand{\ctilde}{\tilde{c}^m_l}
\newcommand{\ctildemod}{\tilde{c}^{|m|}_l}
\newcommand{\chat}{\hat{c}^m_l}
\newcommand{\chatmod}{\hat{c}^{|m|}_l}
\newcommand{\ddx}{\frac{\mathrm{d}}{\mathrm{d}x}}
\newcommand{\pddx}{\frac{\partial}{\partial x}}
\newcommand{\pddy}{\frac{\partial}{\partial y}}
\newcommand{\pddn}{\frac{\partial}{\partial n}}
\newcommand{\dmdxm}{\frac{\mathrm{d}^m}{\mathrm{d}x^m}}

@@

\newcommand{\Pnk}{P_{n,k}}
\newcommand{\Pab}{P^{(a,b)}}
\newcommand{\Pnkab}{P_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\Pmjab}{P_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bigPab}{\bigP^{(a,b)}}
\newcommand{\Dnt}{D^T_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\Pii}{P^{(1,1)}}
\newcommand{\bigPii}{{\mathbb{P}^{(1,1)}}}
\newcommand{\Poo}{P^{(0,0)}}
\newcommand{\bigPoo}{{\mathbb{P}^{(0,0)}}}
\newcommand{\Poi}{P^{(0,1)}}
\newcommand{\Pio}{P^{(1,0)}}
\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}
\newcommand{\laplacewii}{\Delta_W^{(1,1)\to(1,1)}}
\newcommand{\laplacewtt}{\Delta_W^{(2,2)\to(0,0)}}
\newcommand{\laplaceoo}{\Delta^{(0,0)\to(2,2)}}
\newcommand{\biharmonic}{_2\Delta_W^{(2,2)\to(2,2)}}

\newcommand{\element}{\tau}
\newcommand{\refelement}{\hat{\tau}}
\newcommand{\FEset}{\mathcal{T}}
\newcommand{\bigW}{\mathbb{W}}
\newcommand{\bigWab}{\mathbb{W}^{(a,b)}}
\newcommand{\bigWii}{{\mathbb{W}^{(1,1)}}}
\newcommand{\bigQ}{\mathbb{Q}}
\newcommand{\bigQab}{\bigQ^{(a,b)}}

@@

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

We can then obtain the matrix operator for the Laplacian \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\bigWii(x,y) = \Wii(x,y) \: \bigPii(x,y)
\]
to coefficients in the non-weighted space
\[
\bigPii(x,y).
\]
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse, as seen in FIGURE. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$. The matrix operator for the Laplacian we denote $\laplacewii$ acting on the coefficients vector is then given by
\begin{align}
    \laplacewii := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}

Further, we can also obtain the matrix operator for the Laplacian $\Delta$ that will take us from coefficients for expansion in the space
$$
\bigPoo(x,y)
$$
to coefficients in the space
$$
\bigP^{(2,2)}(x,y).
$$
Importantly, again, this operator will have Banded-Block-Banded structure, and hence will be sparse, as seen in FIGURE. We denote this matrix operator that acts on the coefficients vector as $\laplaceoo$, and is given by
\begin{align}
    \laplaceoo := D_x^{(1,1)} \: D_x^{(0,0)} + T^{(1,2)\to(2,2)} \: D_y^{(1,1)} \: T^{(0,1)\to(1,1)} \: D_y^{(0,0)}.
\end{align}

We can use the same technique for constructing a sparse Laplacian operator to construct the Biharmonic operator $\Delta^2$, that ensures we use a basis that satisfies both zero Dirichlet and Neumann conditions. On this note, we can further represent the Laplacian as a map from coefficients in the space 
$$
\bigW^{(2,2)}
$$
to coefficients in the space
$$
\bigPoo.
$$
Note that a function expanded in the $\bigW^{(2,2)}$ basis will satisfy both zero Dirichlet and Neumann boundary conditions on $\Omega$. We denote this matrix operator as $\laplacewtt$, and is given by
\begin{align}
	\laplacewtt := W_x^{(1,1)} \: W_x^{(2,2)} + T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)} \: T_W^{(2,1)\to(1,1)} \: W_y^{(2,2)}.
\end{align}
The sparsity and structure of this matrix can be seen in FIGURE.

The sparsity of the these operators can be shown using a simple integration by parts argument, and can be seen in FIGURE.

@@

\subsection{Calculating non-zero entries of the operator matrices}

First, note that we can expand $\pddx \Pnkab = \sum_{m=0}^{n} \sum_{j=0}^m c_{m,j} \: P^{(a+1, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. By a simple integration by parts argument, we have that the only non-zero coefficients of this expansion are when $m = n-1, n-2$, $j = k, k-2$ and $0 \le j \le m$. These are then calculated by taking inner products, i.e. 
\begin{align*}
c_{m,j} = \frac{\ip<\pddx \Pnkab, P^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}}}{\norm{P^{(a+1, b+1)}_{m,j}}^2_{W^{(a+1, b+1)}}}
\end{align*}
where 
\begin{align}
	&\ip<\pddx \Pnkab, P^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}} \nonumber \\ 
	&= \ip< x \: H_{n-k}^{(a, b+k+\half)} \prime (x), \: \rho(x)^{k+j+2} \: H_{m-j}^{(a, b+1+j+\half)} >_{w_H^{(a, b+\half)}} 
		\: \ip< P_k^{(b,b)}, \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  \nonumber \\ 
	& \quad - k \: \ip< x^2 \: H_{n-k}^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: H_{m-j}^{(a, b+1+j+\half)} >_{w_H^{(a, b+\half)}} 
		\: \ip< P_k^{(b,b)}, \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  \nonumber \\ 
	& \quad + \: \ip< x^2 \: H_{n-k}^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: H_{m-j}^{(a, b+1+j+\half)} >_{w_H^{(a, b+\half)}} 
		\: \ip< y \: P_k^{(b,b)} \prime (y), \: P_j^{(b+1,b+1)} >_{w_P^{(b+1)}}  
\end{align}
We can then use quadrature to calculate the 1D inner products. This is much cheaper than using the 2D quadrature rule.

Further, we can expand $\pddy \Pnkab = \sum_{m=0}^{n} \sum_{j=0}^m c_{m,j} \: P^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. By a simple integration by parts argument, we have that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. This is then calculated by taking inner products, i.e. 
\begin{align*}
c_{n-1,k-1} = \frac{\ip<\pddx \Pnkab, P^{(a, b+1)}_{n-1,k-1} >_{W^{(a, b+1)}}}{\norm{P^{(a, b+1)}_{n-1,k-1}}^2_{W^{(a, b+1)}}}
\end{align*}
where 
\begin{align}
	\ip<\pddx \Pnkab, P^{(a, b+1)}_{n-1,k-1} >_{W^{(a, b+1)}} = \norm{H_0^{(a,b+k+\half)}}^2_{w_H^{(a, b+k+\half)}} 
		\: \ip< P_k^{(b,b)} \prime, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}}
\end{align}
where we note that $\norm{H_n^{(a,b)}}_{w_H^{(a,b)}} = \norm{H_m^{(a,b)}}_{w_H^{(a,b)}}$ $\forall m, n = 0, 1, 2, \dots$.

@@

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigPii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. In matrix-vector notation, our system hence becomes:
\begin{align}
    \laplacewii \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$.
In FIGURE we see the solution to the Poisson equation with zero boundary conditions given in EQN in the half-disk $\Omega$. In FIGURE we show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the Poisson equation with N = 999????. The rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution. \bstodo{Describe what plot tells us}

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.4]{sparsityoflaplacian}
        \label{fig:sparsity}
    	\caption{A "spy" plot of the $\laplacewii$ matrix operator for the Laplacian showing its sparsity.}
	\centering
\end{figure}

\subsection{Inhomogenous variable-coefficient Helmholtz}

Find \(u(x,y)\) given functions $v$, $f : \Omega \to \R$ such that:
\begin{align}
	\begin{cases}
    		\Delta u(x,y) + k^2 v(x,y) \; u(x,y) = f(x,y) \quad \text{in } \Omega \\
		u(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
\end{align}
where $k \in \R$, noting the imposition of zero Dirichlet boundary conditions on $u$.

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigPii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. We can obtain the matrix operator for the variable-coefficient function $v(x,y)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices ${J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T$, yielding an operator matrix of the same dimension as the input Jacobi matries. We can denote the resulting operator acting on coefficients in the $\bigPoo$ space by $V({J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T)$. In matrix-vector notation, our system hence becomes:
\begin{align}
    (\laplacewii + k^2 T^{(0,0)\to(1,1)} \: V({J_x^{(0,0)}}^T, {J_y^{(0,0)}}^T) \: T_W^{(1,1)\to(0,0)}) \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in FIGURE with $v(x,y) = xy^2$ as an example. In FIGURE we see the solution to the inhomogenous variable-coefficient Helmholtz equation with zero boundary conditions given in EQN in the half-disk $\Omega$, with  $v(x,y) = xy^2$ and $f(x,y) = xy e^x$.

We can extend this to constant non-zero boundary conditions by simply noting that the problem 
\begin{align}
	\begin{cases}
    		\Delta u(x,y) + k^2 v(x,y) \; u(x,y) = f(x,y) \quad \text{in } \Omega \\
		u(x,y) = c \in \R \quad \text{on } \partial \Omega
	\end{cases}
\end{align}
is equivalent to letting $u = \tilde{u} + c$ and solving
\begin{align}
	\begin{cases}
    		\Delta \tilde{u}(x,y) + k^2 v(x,y) \; \tilde{u}(x,y) = f(x,y) - k^2 v(x,y) \; =: g(x,y)  \quad \text{in } \Omega \\
		\tilde{u}(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
\end{align}

@@

\subsection{Biharmonic equation}

Find \(u(x,y)\) given a function \(f(x,y)\) such that:
\begin{align}
	\begin{cases}
    		\Delta^2 u(x,y) = 0 \quad \text{in } \Omega \\
		u(x,y) = 0, \quad \frac{\partial u}{\partial n}(x,y) = 0 \quad \text{on } \partial \Omega
	\end{cases}.
\end{align}
where $\Delta^2$ is the Biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. We can construct a matrix operator for $\Delta^2$ that will take coefficients in the space
$$
\bigW^{(2,2)}
$$
to coefficients in the space
$$
\bigP^{(2,2)}.
$$
Note that any function expanded in the $\bigW^{(2,2)}$ basis will satisfy both zero Dirichlet and zero Neumann boundary conditions on $\Omega$. The matrix operator for the Biharmonic operator is then given by
\begin{align}
	\biharmonic = \laplaceoo \: \laplacewtt.
\end{align}
In FIGURE we see the solution to the Biharmonic equation EQN in the half-disk $\Omega$.

\subsection{Heat equation}
Consider
\begin{align}
	\begin{cases}
		\frac{\partial u}{\partial t}(t,x,y) = \Delta u(t,x,y) \quad \text{in } \Omega \\
		u(t, x, y) = 0 \quad \text{on } \partial \Omega \\
		u(0, x, y) = u_0(x,y) := \Wii(x,y) \: f(x,y)
	\end{cases}
\end{align}
noting the imposition of zero Dirichlet boundary conditions.

@@

We can solve this system for the given initial condition $\mathbf{u_0}$ by iterating for $t = 0,1,2,\dots,t_{max}$:
\begin{align}
	&T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} \: (\mathbf{u}_{t+1} - \mathbf{u}_t) = h \: \laplacewii \mathbf{u}_{t+1} \nonumber \\
	\iff& (T^{(0,0)\to(1,1)} \:T_W^{(1,1)\to(0,0)} - h \: \laplacewii) \: \mathbf{u}_{t+1} = T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} \: \mathbf{u}_t \nonumber \\
	\iff& \mathbf{u}_{t+1} = (T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} - h \: \laplacewii)^{-1} T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} \: \mathbf{u}_t \nonumber \\
	\iff& \mathbf{\tilde{u}}_{t+1} = T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} \: (T^{(0,0)\to(1,1)} \: T_W^{(1,1)\to(0,0)} - h \: \laplacewii)^{-1} \: \mathbf{\tilde{u}}_t.
\end{align}

@@

%
\section{Future work - orthogonal polynomials on half-spheres and spherical triangles}
